---
title: "VL Data Exploration"
author: "Morgan Kain"
date: "4/14/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)
library(ggplot2); library(dplyr); library(geobr); library(sf)

## set up ggplot theme
theme_set(theme_bw())
suppressWarnings(
theme_update(axis.text.x = element_text(size = 13),
             axis.text.y = element_text(size = 13),
             axis.title.x = element_text(size = 13),
             axis.title.y = element_text(size = 13),
             legend.title = element_text(size = 12),
             legend.text  = element_text(size = 11),
             panel.grid.major = element_blank(),
             panel.grid.minor = element_blank(),
             strip.background = element_blank(),
             panel.margin = unit(0, "lines"),
             legend.key.size = unit(.45, "cm"),
             legend.key = element_rect(fill = "white"),
             panel.margin.y = unit(0.5, "lines"),
             panel.border = element_rect(colour = "black", 
                                         fill = NA, size = 1),
             strip.text.x = element_text(size = 16, colour = "black", 
                                         face = "bold"))
)

if (!file.exists("VL_cleaned.Rds")) {
## VL data, District mapping data, and a matching of the district names
mun        <- readRDS("Brazil_map.rds")
VL         <- readRDS("prelim_vl_data.rds")
name_match <- read.csv("VL_municipality_match.csv")

## Cleanup of district names between the VL data frame and the map data frame of districts
map_names <- strsplit(as.character(unique(mun$name_muni)), " ") %>% 
  sapply(., FUN = function(x) x) %>% 
  sapply(., FUN = function(x) paste(toupper(substring(x, 1, 1)), substring(x, 2), sep = "", collapse = " "))

VL_names <- strsplit(as.character(VL$NM_MUN), " ") %>% 
  sapply(., FUN = function(x) x) %>% 
  sapply(., FUN = function(x) paste(toupper(substring(x, 1, 1)), substring(x, 2), sep = "", collapse = " "))

VL         <- VL %>% mutate(VL_name = VL_names)
name_match <- name_match %>% dplyr::select(VL_name, Convert_name) %>% rename(munip_name = Convert_name)

VL         <- VL %>% left_join(., name_match)
mun        <- mun %>% rename(munip_name = name_muni)

mun.all   <- mun %>% left_join(., (VL %>% group_by(munip_name) %>% summarize(total_cases = sum(vl_cases))))
mun.month <- mun %>% left_join(., (VL %>% group_by(munip_name, month) %>% summarize(total_cases = sum(vl_cases))))
mun.year  <- mun %>% left_join(., (VL %>% group_by(munip_name, year) %>% summarize(total_cases = sum(vl_cases))))

## Some missing population sizes
pop_size <- VL %>% filter(!is.na(total_population)) %>% group_by(munip_name) %>% 
  slice(1) %>% dplyr::select(munip_name, total_population) %>%
  rename(human_pop = total_population)
  
VL <- VL %>% left_join(., pop_size)

saveRDS(
    list(VL = VL, mun.all = mun.all, mun.month = mun.month, mun.year = mun.year)
  , "VL_cleaned.Rds") 

} else {
all.data  <- readRDS("VL_cleaned.Rds")
VL        <- all.data[[1]]
mun.all   <- all.data[[2]]
mun.month <- all.data[[3]]
mun.year  <- all.data[[4]]
rm(all.data)
}

```

## First pass visualizations of VL cases in Brazil

#### Check distribution, look for weird outliers

The numbers in the upper right of each panel show the total number of cases

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
VL %>% {
  ggplot(., aes(x = vl_cases)) +
    geom_histogram(aes(fill = month, colour = month), bins = 300, position = 'dodge') +
    xlab("Cases") + ylab("Frequency") +
    scale_colour_discrete(name = "Month") +
    scale_fill_discrete(name = "Month") +
    facet_wrap(~year) +
    scale_y_continuous(trans = "pseudo_log", breaks = c(0, 2, 4, 10, 100, 1000, 6000)) +
    geom_text(
      data = (VL %>% group_by(year) %>% summarize(total = sum(vl_cases)))
    , aes(x = 30, y = 1000, label = total)
    )
  }
```

#### Yearly and monthly trends

Individual faded black lines show individual municipalities. The solid red line shows the sum across all municipalities, while the blue line shows the average number of cases across municipalities

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
VL %>% 
  group_by(NM_MUN, year) %>%
  summarize(num_cases = sum(vl_cases)) %>% 
  mutate(year = as.numeric(year)) %>% {
    ggplot(., aes(year, num_cases)) +
      geom_line(aes(group = NM_MUN), alpha = 0.2) +
      scale_y_continuous(trans = "pseudo_log", breaks = c(0, 1, 5, 10, 100, 200, 300, 3000)) +
      geom_line(
        data = (VL %>% group_by(year) %>% summarize(num_cases = sum(vl_cases)) %>% mutate(year = as.numeric(year)))
      , aes(year, num_cases)
      , lwd = 1, colour = "red3"
      ) +
      geom_line(
        data = (VL %>% group_by(year) %>% summarize(num_cases = mean(vl_cases)) %>% mutate(year = as.numeric(year)))
      , aes(year, num_cases)
      , lwd = 1, colour = "blue3"
      ) +
      xlab("Year") + ylab("VL Cases")
  }
```

Individual faded black lines show individual municipalities. The solid red line shows the sum across all municipalities, while the blue line shows the average number of cases across municipalities.

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
VL %>% 
  group_by(NM_MUN, month) %>%
  summarize(num_cases = sum(vl_cases)) %>% 
  mutate(month = as.numeric(month)) %>% {
    ggplot(., aes(month, num_cases)) +
      geom_line(aes(group = NM_MUN), alpha = 0.2) +
      scale_y_continuous(trans = "pseudo_log", breaks = c(0, 1, 5, 10, 100, 200, 300, 3000)) +
      scale_x_continuous(breaks = c(1, 3, 5, 7, 9, 11), labels = c("Jan", "Mar", "May", "Jul", "Sep", "Nov")) +
      geom_line(
        data = (VL %>% group_by(month) %>% summarize(num_cases = sum(vl_cases)) %>% mutate(month = as.numeric(month)))
      , aes(month, num_cases)
      , lwd = 1, colour = "red3"
      ) +
      geom_line(
        data = (VL %>% group_by(month) %>% summarize(num_cases = mean(vl_cases)) %>% mutate(month = as.numeric(month)))
      , aes(month, num_cases)
      , lwd = 1, colour = "blue3"
      ) +
      xlab("Month") + ylab("VL Cases")
  }
```

Look at both month and year trends simultaneously in a heatmap to look for both year and month trends spatially

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
# rand_locs <- unique(VL$munip_name) %>% sample(., 10)
rand_locs <- VL %>% group_by(munip_name) %>% summarize(tot_cases = sum(vl_cases)) %>% 
  arrange(desc(tot_cases)) %>% head(9) %>% dplyr::select(munip_name) %>% unname() %>% unlist()

VL %>% filter(munip_name %in% rand_locs[c(1:6, 8:9)]) %>% {
    ggplot(., aes(as.numeric(year), as.numeric(month))) +
    geom_tile(aes(fill = vl_cases), colour = "white") +
    scale_fill_gradient(low = "grey95", high = "#4472C4"
      , name = "VL Cases") +
    scale_y_reverse(
        breaks = seq(1, 12)
      , labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug","Sep", "Oct", "Nov", "Dec")
      ) + xlab("Year") + ylab("Month") +
    facet_wrap(~munip_name)
  }

```

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
mun.all %>% mutate(munip_highlight = ifelse(munip_name %in% rand_locs, 1, 0)) %>% {
  ggplot(.) + 
  geom_sf(aes(fill  =munip_highlight), color = "#FEBF57", size = 0) +
  scale_fill_gradient(low = "slategray3", high = "firebrick1", name = "Total
Cases") + guides(fill = FALSE)
}
```

#### Over space

Total cases between 2007 and 2018 by municipality

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
ggplot(data = mun.all) + 
  geom_sf(aes(fill = total_cases), color = "#FEBF57", size = 0) +
  scale_fill_gradient(low = "slategray3", high = "firebrick1", name = "Total
Cases")
```

#### Over space and time 

Total cases by year (summed over months). Struggling to get this to print bigger

```{r, fig.width = 12, fig.height = 14}
ggplot(data = 
    (mun.year %>% filter(!is.na(total_cases)))
  ) + 
  geom_sf(aes(fill = total_cases), color = "#FEBF57", size = 0) +
  scale_fill_gradient(low = "slategray3", high = "firebrick1", name = "Total
Cases") + facet_wrap(~year)
```

Total cases by month (summed over years). Struggling to get this to print bigger

```{r, fig.width = 12, fig.height = 14}
ggplot(data = 
    (mun.month %>% filter(!is.na(total_cases)))
  ) +
  geom_sf(aes(fill = total_cases), color = "#FEBF57", size = 0) +
  scale_fill_gradient(low = "slategray3", high = "firebrick1", name = "Total
Cases") + facet_wrap(~month) + 
  theme(axis.text.x = element_text(size = 14, angle = 310, hjust = 0))
```

#### Coarse correlations between predictors and cases

Total Precipitation, each data point is the number of VL cases in a given month, year, and municipality

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}

VL %>% {
  ggplot(., aes(total_precipitation, vl_cases)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Total Precipitation") + ylab("VL Cases (Month)")
}

```

Broken down by municipality to look for variation in the effects of the predictor over space

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
rand_locs <- VL %>% group_by(munip_name) %>% summarize(tot_cases = sum(vl_cases)) %>% 
  arrange(desc(tot_cases)) %>% head(20) %>% dplyr::select(munip_name) %>% unname() %>% unlist()

VL %>% filter(munip_name %in% rand_locs) %>% {
  ggplot(., aes(total_precipitation, vl_cases)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Total Precipitation") + ylab("VL Cases (Month)") +
    facet_wrap(~munip_name)
}

```

Human Population and Municipality Size, each data point is the number of VL cases in a given month, year, and municipality
```{r, fig.width = 9, fig.height = 9, out.width = "100%"}

VL %>% {
  ggplot(., aes(human_pop, vl_cases)) + 
    geom_point(aes(colour = log10(AREA_KM2)), lwd = 3, alpha = 0.5) +
    scale_colour_gradient(name = "log10(KM2)", low = "dodgerblue4", high = "orange") +
    scale_x_log10() +
    xlab("Human Population") + ylab("VL Cases (Month)")
}

```

Median NDVI, each data point is the number of VL cases in a given month, year, and municipality
```{r, fig.width = 9, fig.height = 9, out.width = "100%"}

VL %>% {
  ggplot(., aes(median_ndvi, vl_cases)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Median NDVI") + ylab("VL Cases (Month)")
}

```

Mean Air Temp, each data point is the number of VL cases in a given month, year, and municipality
```{r, fig.width = 9, fig.height = 9, out.width = "100%"}

VL %>% {
  ggplot(., aes(mean_air_temp, vl_cases)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Mean Air Temp") + ylab("VL Cases (Month)")
}

```

Broken down by municipality to look for variation in the effects of the predictor over space

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
rand_locs <- VL %>% group_by(munip_name) %>% summarize(tot_cases = sum(vl_cases)) %>% 
  arrange(desc(tot_cases)) %>% head(20) %>% dplyr::select(munip_name) %>% unname() %>% unlist()

VL %>% filter(munip_name %in% rand_locs) %>% {
  ggplot(., aes(mean_air_temp, vl_cases)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Mean Air Temp") + ylab("VL Cases (Month)") +
    facet_wrap(~munip_name)
}

```

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
VL %>% filter(munip_name %in% rand_locs) %>% {
  ggplot(., aes(month, mean_air_temp)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Month") + ylab("Mean Air Temp") +
    facet_wrap(~munip_name)
}

```

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
VL %>% filter(munip_name %in% rand_locs) %>% {
  ggplot(., aes(month, vl_cases)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Month") + ylab("VL Cases") +
    facet_wrap(~munip_name)
}

```

GDP over time and correlation to cases

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
rand_locs <- VL %>% group_by(munip_name) %>% summarize(tot_cases = sum(vl_cases)) %>% 
  arrange(desc(tot_cases)) %>% head(20) %>% dplyr::select(munip_name) %>% unname() %>% unlist()

VL %>% {
  ggplot(., aes(log10(GDP_1000s_R), vl_cases)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("GDP") + ylab("VL Cases (Month)")
}

```

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
VL %>% filter(munip_name %in% rand_locs) %>% {
  ggplot(., aes(year, log10(GDP_1000s_R))) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Year") + ylab("GDP") +
    facet_wrap(~munip_name)
}
```

```{r, fig.width = 9, fig.height = 9, out.width = "100%"}
VL %>% filter(munip_name %in% rand_locs) %>% {
  ggplot(., aes(year, vl_cases)) + 
    geom_point(lwd = 2, alpha = 0.5) +
    xlab("Year") + ylab("VL Cases") +
    theme(axis.text.x     = element_text(angle = 300, hjust = 0, colour = "grey50", size = 9)) +
    facet_wrap(~munip_name)
}
```

### Some notes:
 
##### In most municipalities seasonality is minimal
##### In most municipalities there is relatively little trend over years (some are decreasing, some are increasing?)
##### Is there a discernable pattern in cases over space over years (e.g., movement westward or...)?
##### Is there a discernable pattern in cases over months across space?
##### If using a random effect framework, how much goes into the random effects? (1 | munip) or much more (1 + year + beta | munip) ?

